---
title: "Retrieving NSAP Data with R"
subtitle: "SPC 2026"
author: "Jessica Leiria Schattschneider"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
---

# Introduction

This document help NSAP users to retrieve fishing data from NSAP reports. The goal 
is to see how R can help you access and filter data that matters to your work.

**Steps:**

1.  Connect to NSAP - you need to have a password for NSAP
2.  See what reports are available
3.  Filter reports to find what we need
4.  Download the data

------------------------------------------------------------------------

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# Clear the workspace - start fresh!
rm(list=ls())

# Load the libraries we need
library("DT")        # For interactive tables
library("dplyr")     # For data manipulation
library("stringr")   # For working with text
library("purrr")     # For data operations
library(httr)

base_url = "https://nsap.ofp.spc.int/" # Base URL for T2 API (https://www.spc.int) 
                                       # or to nsap (https://nsap.ofp.spc.int/)
# Load custom functions from utils.R
source("utils.R")

dir.create("./data/", showWarnings = FALSE, recursive = TRUE)

```


# Step 1: Connect to NSAP

To access data, we need to identify ourselves. This is like logging into a website.

But first, you will need to make sure you have (manually) created a file called `.env`
in the same path where this script is stored. After you create it, copy the content from 
file `envtemplate` and paste in `.env`, and add your NSAP password after `NSAP_PASSWORD`. 
Make sure there is an empty line after at the end of the document.

You can also add your Tufman 2 password after `TUF_PASSWORD` if you have one and want to
download data from this platform too, but this is not required. If you are only interested
in NSAP data, you can leave the `TUF_PASSWORD` line as is or delete the line entirely.

You will also need to include the email associated with your NSAP credentials in the 
`user_name` variable in the chunck below.

```{r authentication}
# Enter your details here

user_name = "YOUR-EMAIL@ORG.INT" # Your username (NSAP login credential)
region_code = "R03"             # Select your region of interest

# Load or create your access token
token <- load_token(user_name, country_code = 'HQ', base_url) 

```


------------------------------------------------------------------------

# Step 2: What Reports Are Available?

Let's see all the reports you have access to in NSAP.

```{r load-reports, message=FALSE, warning=FALSE, echo=FALSE, include = FALSE}
# Get list of all available reports
existing_reports <- get_list_of_t2_reports(
  token, 
  region_code, 
  user_name, 
  base_url = 'https://nsap.ofp.spc.int/',
  replace = FALSE # talk about what this is doing
) |>
  data.frame()

result <- httr::GET(
  url = "https://nsapapi.ofp.spc.int/api/ReportDefinition/AllSimple",
  add_headers(
    "accept"        = "application/json, text/plain, */*",
    "authorization" = paste("Bearer", token),
    "content-type"  = "application/json",
    "tufinstance"   = "R03",
    "tufmodule"     = "General",
    "tufuser"       = "jessicals@spc.int"
  )
)

# status_code(result)
# How many reports do we have access to?

```

Now let's look at them in an interactive table:

```{r display-reports, echo=FALSE}

cat("Total reports available:", nrow(existing_reports), "\n")


# Create an interactive table
DT::datatable(
  existing_reports |>
    dplyr::select(
      Id = user_report_id,
      Filters = option_labels,
      Title = title, 
      Description = description, 
      `Last modified` = last_modified_date_time
    ) |>
    arrange(Title),
  filter = "top",  # Adds search boxes at the top
  options = list(
    pageLength = 10,
    dom = 'lftrip',
    initComplete = JS(
      "function(settings, json) {",
      "$(this.api().table().header()).css({'background-color': '#000', 'color': '#fff'});",
      "}"
    )
  )
)
```


------------------------------------------------------------------------

# Step 3: Filtering by Specific IDs

Sometimes we know exactly which reports we want by their ID numbers:

```{r filter-reports-v2}

# update report_ids based on the reports you are interested in
report_ids = c(7)

# Select specific reports by ID
filtered_reports <- existing_reports |>
  filter(user_report_id %in% report_ids)

cat("Selected reports:", nrow(filtered_reports), "\n")
```



## Step 3.1: Explore Report Attributes

Before downloading data, let's see what attributes (filters/parameters) are available:

```{r explore-attributes}
# Extract all unique attributes from our filtered reports
filtered_attrs <- filtered_reports$report_attrs |>
  paste(collapse = ",") |>           # Combine all attributes
  str_split(",") |>                  # Split into individual items
  unlist() |>                        # Flatten the list
  str_trim() |>                      # Remove extra spaces
  discard(~ .x == "") |>             # Remove empty entries
  unique() |>                        # Keep only unique values
  sort()                             # Sort alphabetically

# Display the attributes
cat("Available attributes:\n")
print(filtered_attrs)
```



## Step 3.2: Define Your Filter Values


Make sure you check in NSAP the attributes that are mandatory, unfortunately we cannot
access this information programatically. 

Now we tell R what specific data we want based on the available attributes we will need to inform for the current selection 
of reports:

```{r define-parameters}
attrs <- list(
  "year" = 2024
)

```

You're requesting data for:
```{r define-parameters-print, echo = FALSE}

for (i in 1:length(attrs)){
  
  cat(paste0(
    names(attrs[i]),
    ": ", attrs[[i]], " \n"
  )
  )

}

```

------------------------------------------------------------------------

# Step 6: Download the Data!

Finally, we're ready to download our data:

```{r download-data, eval=FALSE}
# Get the report data based on our filters
report_data <- get_reports(
  token = token, 
  user_name = user_name,
  base_url = "https://nsapapi.ofp.spc.int/api/ReportDefinition/DownloadResults",
  country_code = region_code, 
  filtered_reports = filtered_reports, 
  attrs = attrs,
  record_current_date = FALSE,
  overwrite = TRUE
)

# Preview the data structure
cat("Data downloaded successfully!\n")
```
