---
title: "File Validation Report"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    embed-resources: true
    code-fold: true
params:
  last_run: "2024-04-04"
---
```{r setup, include=FALSE}
library(tidyverse)
library(gt)
library(scales)
library(janitor)

data <- read.csv("output/log_validation.csv")

truncate_str <- function(x, n = 50) {
  if_else(nchar(x) > n, paste0(substr(x, 1, n), "..."), x)
}

data <- data |>
  mutate(
  errors.check        = truncate_str(errors.check, 50),
  errors.failure_case = truncate_str(errors.failure_case, 80)
) |>
  mutate(errors.check = case_when(errors.check == "column_in_dataframe" ~ "missing_column_in_dataframe",
                   .default = errors.check))  |>
  mutate( errors.column = case_when(is.na(errors.column) ~ errors.failure_case,
                                    .default = errors.column))


n_total <- data |> distinct(filename) |> nrow()
n_passed <- data |> distinct(filename, status) |> filter(status != "Validation failed") |> nrow()
n_failed <- n_total - n_passed
pct_passed <- round(100 * n_passed / n_total, 1)

res_filename_summary <- data |>
  distinct(status, filename) |>
  group_by(status) |>
  summarise(n = n()) |>
  mutate(percent = round(100 * n / sum(n), 2)) |>
  ungroup()

res_filename_summary_cc <- data |>
  distinct(status, filename, country_code) |>
  group_by(country_code, status) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(percent = round(100 * n / sum(n), 2)) |>
  ungroup()

errors_check <- data |>
  filter(status == "Validation failed") |>
  group_by(country_code, filename, errors.column, errors.check, errors.failure_case) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(country_code, filename)

error_summary_by_check <- data |>
  filter(status == "Validation failed") |>
  group_by(errors.check, errors.column) |>
  summarise(
    n_failures  = n(),
    n_files     = n_distinct(filename),
    n_countries = n_distinct(country_code),
    .groups = "drop"
  ) |>
  arrange(desc(n_failures))
```

## Overview

- Validation run: **`r params$last_run`**

- Total files: **`r n_total`**

- Passed: **`r n_passed` (`r pct_passed`%)**

- Failed: **`r n_failed`**

---

## By country

Files that failed validation per country. Countries with no failures are omitted.
```{r country-table, echo=FALSE}
res_filename_summary_cc |>
  pivot_wider(names_from = status, values_from = c(n, percent), values_fill = 0) |>
  clean_names() |>
  select(country_code,
         passed = `n_all_data_passed_validation`,
         pct_passed = `percent_all_data_passed_validation`,
         failed = `n_validation_failed`,
         pct_failed = `percent_validation_failed`) |>
  filter(failed > 0) |>
  arrange(desc(pct_failed)) |>
  gt() |>
  cols_label(
    country_code = "Country",
    passed       = "Passed",
    pct_passed   = "%",
    failed       = "Failed",
    pct_failed   = "%"
  ) |>
  fmt_number(columns = c(pct_passed, pct_failed), decimals = 1) |>
  data_color(
    columns = pct_failed,
    palette = c("#FDECEA", "#C0392B")
  ) |>
  tab_spanner(label = "Passed", columns = c(passed, pct_passed)) |>
  tab_spanner(label = "Failed", columns = c(failed, pct_failed)) |>
  tab_options(table.width = pct(100), table.font.size = 13)
```

---

## Error summary

The table below identifies which validation checks and columns are generating the most failures, across all countries. These are the highest-priority issues to resolve.
```{r error-summary-table, echo=FALSE}
error_summary_by_check |>
  gt() |>
  cols_label(
    errors.check        = "Check",
    errors.column       = "Column",
    n_failures          = "Total failures",
    n_files             = "Files affected",
    n_countries         = "Countries"
  ) |>
  data_color(
    columns = n_failures,
    palette = c("#FDECEA", "#C0392B")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = errors.check)
  ) |>
  tab_options(table.width = pct(100), table.font.size = 13)
```

---

## Error detail by country

Expand each country to review the specific files and failure cases that need attention.
```{r error-detail, results='asis', echo=FALSE}
countries_with_errors <- errors_check |> pull(country_code) |> unique() |> sort()

for (cc in countries_with_errors) {
  cat(paste0("\n### ", cc, "\n\n"))
  
  tbl <- errors_check |>
    filter(country_code == cc) |>
    select(filename, errors.column, errors.check, errors.failure_case, n) |>
    arrange(filename, errors.check)
  
  print(
    tbl |>
      gt() |>
      cols_label(
        filename             = "File",
        errors.column        = "Column",
        errors.check         = "Check",
        errors.failure_case  = "Failure case",
        n                    = "Count"
      ) |>
      data_color(columns = n, palette = c("#FFF8E1", "#E65100")) |>
      tab_options(table.width = pct(100), table.font.size = 12)
  )
  
  cat("\n\n")
}
```