---
title: "SPC Coastal Fisheries Web Application - API"
subtitle: "R Implementation"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
author: "Jessica LS"
date: today
---

# Overview

This notebook demonstrates how to interact with the **SPC (Pacific Community) Coastal Fisheries Web Application** API to retrieve and analyze landing survey data using R.

1. **Authenticate** with the SPC Coastal Fisheries API
2. **Explore** available authorities, surveys, and species
3. **Query** landing survey reports
4. **Export** data in JSON and CSV formats
5. **Filter** data by specific criteria

## Prerequisites

Before running this notebook, ensure you have:

- Created a `.env` file with your SPC portal credentials
- Installed required R packages (see below)
- Valid access to the SPC Coastal Fisheries Web Application

> **Note:** If you don't have credentials yet, contact Franck Magron, Olivier Loyau, or the Coastal Fisheries team to request access to the [SPC Coastal Fisheries Portal](https://www.spc.int/CoastalFisheries/).

## Installation

```r
# Install required packages
install.packages(c("httr", "jsonlite", "dplyr", "readr", "dotenv"))
```

---

# Import Libraries and Setup Environment

First, we'll import the necessary R libraries and load our environment variables containing the API credentials.

```{r}
#| label: setup
#| message: false
#| warning: false

# Load required libraries
library(httr)      # For making HTTP API calls
library(jsonlite)  # For parsing JSON responses
library(dplyr)     # For data manipulation
library(readr)     # For reading/writing CSV files
library(dotenv)    # For loading environment variables from .env file

# Load environment variables from .env file
load_dot_env()

# Get credentials from environment variables
baseUrl <- Sys.getenv("BASEURL")
```

---

# Step 1: Authentication

Now we'll authenticate with the SPC Coastal Fisheries API using the `/account/SignIn` endpoint. This establishes a session and provides cookies for subsequent API calls.

```{r}
#| label: authentication

# Authenticate with the SPC Coastal Fisheries API
connection <- POST(
  paste0(baseUrl, "account/SignIn"),
  body = list(
    username = Sys.getenv("USER"),
    password = Sys.getenv("PASSWORD")
  ),
  encode = "form"
)

# Check if authentication was successful
if (status_code(connection) == 200 && length(content(connection)) > 0) {
  cat("✅ Authentication successful!\n")
  
  # Store session cookies for reuse in other API calls
  session_cookies <- cookies(connection)
  cat("Session cookies stored for future API requests\n")
} else {
  cat("❌ Authentication failed!\n")
  cat("Please check your credentials in the .env file\n")
  cat(paste("Status Code:", status_code(connection), "\n"))
  stop("Authentication failed")
}
```

---

# Step 2: Explore Available Authorities

Authorities represent different regions or organizations that manage coastal fisheries. Let's see which authorities your account can access.

```{r}
#| label: authorities

# Get authorities
authorities_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/GetAuthorities"),
  body = list(),
  encode = "form"
)

# Parse JSON response
authorities <- content(authorities_response, "parsed")

# Convert to data frame for easier analysis
authorities_df <- bind_rows(authorities)

cat(paste("📊 Response: Retrieved", length(authorities), "authorities\n\n"))
cat("First few authorities:\n")
print(head(authorities_df))
```

```{r}
#| label: unique-authorities

# Analyze the unique authorities available
unique_auths <- unique(authorities_df$Name)

cat(paste("\n🌍 Your account has access to", length(unique_auths), "unique authorities:\n"))
cat(paste(rep("=", 60), collapse = ""), "\n")

# Display authorities in a formatted way
for (i in seq_along(unique_auths)) {
  cat(sprintf("%2d. %s\n", i, unique_auths[i]))
}

cat(paste(rep("=", 60), collapse = ""), "\n")

# Store the first authority for use in subsequent examples
first_authority <- if (length(unique_auths) > 0) unique_auths[1] else NULL

if (!is.null(first_authority)) {
  cat(paste("📌 We'll use '", first_authority, "' as our example authority for the following demonstrations\n", sep = ""))
} else {
  cat("⚠️ No authorities found for this account\n")
}
```

---

# Step 3: Discover Available Surveys

Explore what surveys are available for the example authority (above).

```{r}
#| label: surveys

# Get surveys for the first authority
cat(paste("Fetching surveys for authority: '", first_authority, "'\n", sep = ""))
cat("API Call: POST /FieldSurveys/LdsStatistics/GetSurveys\n")

surveys_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/GetSurveys"),
  body = list(authorities = list(first_authority)),  # Filter by specific authority
  encode = "json"
              )

# Parse JSON response
surveys <- content(surveys_response, "parsed")

# Convert to data frame for analysis
surveys_df <- bind_rows(surveys)
unique_surveys <- unique(surveys_df$Name)

cat(paste("\n📈 Found", length(unique_surveys), "surveys for '", first_authority, "':\n", sep = ""))
cat(paste(rep("=", 60), collapse = ""), "\n")

# Display surveys in a formatted way
for (i in seq_along(unique_surveys)) {
  cat(sprintf("%2d. %s\n", i, unique_surveys[i]))
}

cat(paste(rep("=", 60), collapse = ""), "\n")
```

---

# Step 4: Explore Available Species

Let's discover what fish species are recorded in the surveys for our example authority.

```{r}
#| label: species

# Get species information for the first authority
cat(paste("Fetching species data for authority: '", first_authority, "'\n", sep = ""))

species_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/GetSpecies"),
  body = list(authorities = list(first_authority)),  # Filter by specific authority
  encode = "json"
              )


# Parse JSON response
species <- content(species_response, "parsed")

# Convert to data frame for analysis
species_df <- bind_rows(species)

# Analyze species groups and individual species
unique_species_groups <- unique(species_df$SpeciesGroup)
unique_species <- unique(species_df$Species)

cat(paste("\n🐟 Found", length(unique_species), "unique species organized into", 
          length(unique_species_groups), "species groups\n"))
cat("\nSpecies Groups:\n")
cat(paste(rep("=", 50), collapse = ""), "\n")

for (i in seq_along(unique_species_groups)) {
  group <- unique_species_groups[i]
  species_in_group <- species_df %>% 
    filter(SpeciesGroup == group) %>% 
    pull(Species) %>% 
    n_distinct()
  cat(sprintf("%2d. %s (%d species)\n", i, group, species_in_group))
}

cat(paste(rep("=", 50), collapse = ""), "\n")
```

---

# Step 5: Discover Available Reports

Reports are predefined data structures that organize the survey data in different ways. Let's see what reports are available.

```{r}
#| label: reports

# Get list of available reports
cat("Fetching available landing survey reports...\n")
cat("API Call: POST /FieldSurveys/LdsStatistics/GetReports\n")

reports_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/GetReports"),
  body = list(),
  encode = "form"
)

# Parse JSON response
reports <- content(reports_response, "parsed")

# Convert to data frame for analysis
reports_df <- bind_rows(reports)

cat(paste("\n📋 Found", nrow(reports_df), "available reports:\n"))
cat(paste(rep("=", 80), collapse = ""), "\n")

# Display reports with their IDs for reference
for (i in 1:nrow(reports_df)) {
  cat(sprintf("%2d. %s\n", i, reports_df$Name_EN[i]))
}

cat(paste(rep("=", 80), collapse = ""), "\n")

# Store the first report for demonstration
first_report <- if (length(reports) > 0) reports[[1]] else NULL
```

---

# Step 6: Export Report Data as JSON

Now let's export actual survey data from our example report. We'll start with the JSON format, which is great for programmatic analysis.

```{r}
#| label: export-json

# Export report data as JSON for analysis
firstReportId <- reports[[1]]$Id
report_name <- reports[[1]]$Name_EN

cat(paste("Exporting data from report: '", report_name, "'\n", sep = ""))
cat(paste("Report ID:", firstReportId, "\n"))
cat("API Call: POST /FieldSurveys/LdsStatistics/ExportDataAsJson\n")

# Request data export
json_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/ExportDataAsJson"),
  body = list(
    authorities = list(),      # Empty = all authorities (optional)
    reportId = firstReportId,  # Mandatory: which report to export
    surveyIds = list(),        # Empty = all surveys (optional)
    speciesIds = list()        # Empty = all species (optional)
  ),
  encode = "json",
)

# Parse the JSON response
response_data <- content(json_response, "parsed")

cat("\n✅ Data export successful!\n")

# Convert the data to a data frame for analysis
if ("data" %in% names(response_data)) {
  df_result <- bind_rows(response_data$data)
  
  cat(paste("Retrieved", nrow(df_result), "data records\n"))
  cat(paste("Data contains", ncol(df_result), "columns:\n"))
  cat(paste(names(df_result), collapse = ", "), "\n")
  
  cat(paste("\n📊 First 5 records from '", report_name, "':\n", sep = ""))
  cat(paste(rep("=", 100), collapse = ""), "\n")
  print(head(df_result, 5))
  cat(paste(rep("=", 100), collapse = ""), "\n")
} else {
  cat("⚠️ No 'data' field found in response\n")
}
```

---

# Step 7: Export Report Data as CSV

CSV format is perfect for sharing with colleagues or importing into Excel and other tools. Let's export the same data as a CSV file.

```{r}
#| label: export-csv

# Export the same report data as CSV
cat(paste("Exporting '", report_name, "' data as CSV file...\n", sep = ""))
cat("API Call: POST /FieldSurveys/LdsStatistics/ExportDataAsCsv\n")

csv_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/ExportDataAsCsv"),
  body = list(
    authorities = list(),          # Empty = all authorities (optional)
    reportId = firstReportId,      # Mandatory: which report to export
    surveyIds = list(),            # Empty = all surveys (optional)
    speciesIds = list()            # Empty = all species (optional)
  ),
  encode = "json"
)

# Create output directory if it doesn't exist
if (!dir.exists("output_data")) {
  dir.create("output_data")
}

# Save the CSV data to file
output_filename <- "output_data/catchdata.csv"
csv_content <- content(csv_response, "text", encoding = "UTF-8")
write_file(csv_content, output_filename)

cat("\n✅ CSV export successful!\n")
cat(paste("📁 Data saved to:", output_filename, "\n"))

# Show some statistics about the exported file
file_size <- file.info(output_filename)$size

# Count lines in CSV (approximate record count)
line_count <- length(readLines(output_filename, warn = FALSE))

cat(paste("📈 Approximate records:", format(line_count - 1, big.mark = ","), "(excluding header)\n"))
cat(paste("📂 File location:", normalizePath(output_filename), "\n"))
```

---

# Step 8: Advanced Filtering - Export Specific Data

The API allows you to filter data by authorities, surveys, and species. This is useful when you need specific subsets of data rather than everything.

```{r}
#| label: filtered-export

# Example of filtered data export - specific authority and species
cat("🔍 Demonstrating filtered data export...\n\n")

# Use a different report for variety (if available)
if (length(reports) > 1) {
  second_report <- reports[[2]]
  selected_report_id <- second_report$Id
  selected_report_name <- second_report$Name_EN
} else {
  # Use the first report if only one is available
  selected_report_id <- firstReportId
  selected_report_name <- report_name
}

cat(paste("📋 Using report: '", selected_report_name, "'\n", sep = ""))
cat(paste("🌍 Filtering by authority: '", first_authority, "'\n", sep = ""))

# Get a specific species ID for filtering (use first available species)
if (nrow(species_df) > 0) {
  sample_species <- species_df[1, ]
  species_filter <- list(species_df$Id[1])
  species_name <- species_df$Species[1]
  cat(paste("🐟 Filtering by species: '", species_name, "'\n", sep = ""))
} else {
  # Use example species ID
  species_filter <- list("d791be4d-593f-42f0-9a97-8c8d2181b9fe")
  cat("🐟 Using example species filter\n")
}

cat("\n🔧 API Call: POST /FieldSurveys/LdsStatistics/ExportDataAsCsv (with filters)\n")

# Make filtered API request
filtered_response <- POST(
  paste0(baseUrl, "FieldSurveys/LdsStatistics/ExportDataAsCsv"),
  body = list(
    authorities = list(first_authority),    # Filter by specific authority
    reportId = selected_report_id,          # Mandatory report ID
    surveyIds = list(),                     # Could filter by specific surveys
    speciesIds = species_filter             # Filter by specific species
  ),
  encode = "json"
)

# Create output directory if it doesn't exist
if (!dir.exists("output_data")) {
  dir.create("output_data")
}

# Save filtered results
filtered_filename <- "output_data/catchdata_specific_species.csv"
filtered_content <- content(filtered_response, "text", encoding = "UTF-8")
write_file(filtered_content, filtered_filename)

cat("\n✅ CSV export successful!\n")
cat(paste("📁 Data saved to:", filtered_filename, "\n"))

# Show some statistics
filtered_line_count <- length(readLines(filtered_filename, warn = FALSE))
cat(paste("📈 Approximate records:", format(filtered_line_count - 1, big.mark = ","), "(excluding header)\n"))
cat(paste("📂 File location:", normalizePath(filtered_filename), "\n"))
```

---

**Done!**

You can now:

1. **Analyze the data** using R's powerful data analysis tools (dplyr, ggplot2, etc.)
2. **Modify filters** to get different subsets (change authorities, surveys, species)
3. **Integrate** this code into your own data processing workflows
4. **Export** data in formats suitable for sharing with colleagues
